### 2020-08-11 paper.tex
Model Stuff

The Observer design pattern\cite{gamma1995design} facilitates the implementation of RAs. Variations of the pattern are omnipresent today and can be found in places like the Document Object Model (\emph{DOM})\cite{alabor:2019:reactiveappllications}. With Reactive Programming (\emph{RP}), a programming paradigm rooted in functional programming (\emph{FP}), software engineers have yet another tool to choose from to implement a RA. With RP, a domain-specific language (\emph{DSL}) describes \emph{how} data flows through an application, how it depends on each other and how it needs to be transformed over time declaratively. A runtime environment then takes care to fulfill and execute the declared data flows\cite{10.1145/2501654.2501666}.

According to the IEEE Standard Glossary of Software Engineering, \emph{debug} is an activity ``to detect, locate, and correct faults in a computer program.''\cite{ieeeglossary} From interpreting memory dumps, manually adding print statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, debugging utilities took different forms over time.

@@

These exemplary questions demonstrate the limits of a traditional controlflow focused debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow specific inquiry. Regardless of previous research efforts \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} in the field of RP debugging, we will show in this paper that engineers struggle nonetheless in applying effective techniques and utilities. For doing so, we interviewed several software engineers and collected ``war stories'' about the challenges they face in their day-to-day jobs in applying RP. Based on this collected evidence, we will validate these statements in an observational study using RxJS and search for an answer to our first research question:

@@

In response to this, we are going to present concepts on how to resolve previously identified problems and answer the second research question:

@@

The future implementation and validation of these proposals lead to our third research question which will be part of future work, not part of this paper:

@@

We will conclude this introductory section with the clarification of important terms. Section~\ref{sec:interviews} presents the interview results along with the collected war story reports. After we validate them in an observational study in Section~\ref{sec:study}, we will have a look at current and previous research regarding the debugging of RAs in Section~\ref{sec:discussion}. We will finally answer RQ2 in Section~\ref{sec:future} ``Future Work''.

@@

\subsection{Debugging Model and Concepts}

Where the standard glossary defines \emph{debug} as an activity ``to detect, locate, and correct faults in a computer program''\cite{ieeeglossary} in general, Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} formalized the inner workings of the debugging process with the Debugging Process Model further:

@@

TODO LEAD OVER

We divide debugging concepts roughly into three categories: Traditional (i) \emph{imperative-focused debuggers} provide necessary functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, we gain access to the current call stack and the values assigned to variables of a given frame. Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''

Where traditional debuggers can inspect individual transformations of an RP based program, they fall short in understanding the underlying model of the dataflow graph. RP provides its challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g., the predecessor transformation according to the dataflow graph. Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph. A (ii) \emph{reactive debugger} knows about the internal model of an RP runtime system. It leverages on it and provides specialized tools e.g., to navigate or visualize the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.

Traditional, as well as reactive debuggers work with the \emph{current} state of a program's execution only. They lack knowledge about what happened before or what is going to happen in the future. This shortcoming is tedious, especially when debugging a problem depending on many complex circumstances in a system. An (iii) \emph{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly. Instead, it records runtime telemetry and provides an interface for later inspection. Engineers can ``time travel'' back and forth through the program execution trace without the hassle of reconstructing a given failure situation over and over again. This concept comes with its own set of challenges though: Once decided \emph{what} information should be recorded (everything, only specific parts of the program, one thread, multiple threads, \dots), demands on storage performance and capacity increase while the pool of collected data grows. As a result, user experience (\emph{UX}) concepts need to be aligned so engineers can interact with the sheer amount of information efficiently\cite{5287015}.

@@

``Reactive Extensions'' (\emph{ReactiveX}) is an open-source project. Its members and contributors defined a generic description of an RP API. They further provide reference implementations of this API along with RP language extensions for various programming languages like Java, C\#, and JavaScript\footnote{\url{http://reactivex.io/languages.html}}. ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}. The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables. It can push (``emit'') values to its dependents. Multiple depending Observables form and describe a dataflow graph eventually. A multitude of operators allows the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable. This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables. Values emitted from an Observable can be subscribed to at any point, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.

RxJS is the reference implementation of the ReactiveX API specification for JavaScript and TypeScript. It is used thoroughly by large projects like Angular\cite{angualrrxjs}. Listing \ref{lst:rp-with-rxjs} shows an example of RP using RxJS in TypeScript.

@@

		of(0, 1, 2, 3, 4).pipe(	// Create Observable with Integers
			filter(i => i < 4),		// Omit Integers >= 4
			map(i => i * 2)				// Multiply Integer with 2
		).subscribe(console.log) // Emits 0, 2, 4, 6
	\end{minted}
	\caption{Basic RxJS example creating an Observable emitting five integers. Each integer is processed by two operators and finally written to the console.}
	\label{lst:rp-with-rxjs}
\end{listing}

@@

\begin{figure}[H]
	\centering
	\tikzfig{marble-diagram}
	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rp-with-rxjs}. From left to right, each marble represents an emitted value. The vertical line at the last marble indicates that the Observable completed.}
	\label{fig:marble-diagram}
\end{figure}

>>>>

The Observer design pattern\cite{gamma1995design} facilitates the implementation of RAs. Variations of the pattern are omnipresent today and can be found in places like the Document Object Model (\emph{DOM})\cite{alabor:2019:reactiveappllications}. With Reactive Programming (\emph{RP}), a programming paradigm rooted in functional programming (\emph{FP}), software engineers have yet another tool to choose from to implement a RA. A domain-specific language (\emph{DSL}) describes \emph{how} data flows through an application, how it depends on each other and how it needs to be transformed over time declaratively. A RP runtime environment then takes care to fulfill and execute the declared data flows\cite{10.1145/2501654.2501666}.

According to the IEEE Standard Glossary of Software Engineering, \emph{debug} is an activity ``to detect, locate, and correct faults in a computer program.''\cite{ieeeglossary} From interpreting memory dumps, manually adding log statements to trace program execution up to the point where specialized debugging programs can interrupt a running process and interact with it on a low level, debugging utilities took different forms over time.

@@

These exemplary questions demonstrate the limits of a traditional controlflow oriented debugger which is not able to interpret RP abstractions, and as a result, cannot give the correct answer to a dataflow specific inquiry. Regardless of previous research efforts \cite{10.1145/2577080.2577083} \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} in the field of RP debugging, we will show in this paper that engineers struggle nonetheless in applying effective techniques and utilities. For doing so, we interviewed several software engineers and collected ``war stories'' about the challenges they face in their day-to-day jobs when applying RP. Based on this collected evidence, we will validate their statements in an observational study using RxJS and search for an answer to our first research question:

@@

In response to this, we are going to present concepts on how to resolve previously identified challenges and answer the second research question:

@@

The implementation and validation of these proposals lead to our third and last research question, whose answer will be part of a future iteration on the topic:

@@

We will conclude this introductory section with the clarification of important terms. Section~\ref{sec:interviews} continues with insights on the conducted interviews and the collected war story reports. We present our observational study intended to validate former insights in Section~\ref{sec:study} and have a look at previous research efforts regarding the debugging of RAs in Section~\ref{sec:discussion}. Finally, we will answer RQ2 in Section~\ref{sec:future} ``Future Work''.

@@

\subsection{Debugging Process Model}

Where the standard glossary defines \emph{debug} as an activity ``to detect, locate, and correct faults in a computer program''\cite{ieeeglossary} in general, Layman et al. \cite{Layman_Diep_Nagappan_Singer_Deline_Venolia_2013} formalized debugging as an iterative hypothesis refinement process, as shown in Figure~\ref{fig:debugging-process-model}:

@@

The process consists of three steps and includes a feedback loop: After the engineer (i) gathered sufficient context information (e.g. ways to reproduce the failure or details about external factors) and understands the situation satisfactory, they generate a hypothesis on the origins of the bug or what impact a change made to the program might have. With the intent to proof their hypothesis, the engineer then (ii) instruments the defective program using suitable tools (e.g. adding log statements, setting breakpoints or removing code parts). Finally, the instrumented system gets (iii) challenged against the formed hypothesis. E.g., code statements are executed step by step using a debugger or trace logs are analyzed and compared against expected behavior. If the hypothesis turns out to be correct, the debugging process stops. If not, the newly gained knowledge about the problem is used to build a refined hypothesis and start a new iteration.

\subsection{Debugger Concepts}

Software engineers have an arsenal of tools and utilities at hand which help them to interact with and gain insight on the behavior of a defective program. Tools range from the instrumentation of source code with trace log statements (manually or automated) to specialized utilities allowing them to directly interact with a program at runtime.

We will use three main concepts to categorize debugging utilities in this paper: Traditional (i) \emph{imperative-focused debuggers} provide functionality to interact with programs at runtime: Once a breakpoint pauses the program execution, they provide access to the current call stack and the values assigned to variables of a given stack frame. Manual control of the program execution allows inspecting its behavior step by step as well as assigning new values to variables ``on-the-fly.''

RP provides its challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g., the predecessor transformation according to the dataflow graph. Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph. A (ii) \emph{reactive debugger} can interpret the underlying graph model of a RP runtime system. It leverages on it and provides specialized tools e.g., to navigate, visualize or instrument the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}.

Traditional, as well as reactive debuggers work with the \emph{current} state of a programs execution only. They lack information about what happened before or what is going to happen in the future. This shortcoming is tedious, especially when debugging a problem depending on many complex circumstances in a system. An (iii) \emph{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly. Instead, it records runtime telemetry and provides an interface for later inspection. Engineers can ``time travel'' back and forth through the program execution trace without the hassle of reconstructing a given failure situation over and over again.

@@

``Reactive Extensions'' (\emph{ReactiveX}) is an open-source project. Its members and contributors created a generic description of a RP API. They further provide reference implementations of this API along with RP language extensions for various programming languages like Java, C\#, or JavaScript\footnote{\url{http://reactivex.io/languages.html}}. ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}. The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables. It can push (``emit'') values to its dependents. Multiple depending Observables form and describe a dataflow graph eventually. A multitude of operators allows the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable. This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables. Values emitted from an Observable can be subscribed, which is closely related to the Observer design patterns \mintinline{Java}{attach} method.

RxJS is the reference implementation of the ReactiveX API specification for JavaScript. Its current major version 6 is implemented using TypeScript and is used thoroughly by large projects like Angular\cite{angualrrxjs}. Listing \ref{lst:rp-with-rxjs} shows an example of RP using RxJS in TypeScript.

@@

		of(0, 1, 2, 3, 4).pipe(	// Create Observable with Integers
			filter(i => i < 4),		// Omit Integers >= 4
			map(i => i * 2)				// Multiply Integer with 2
		).subscribe(console.log) // Emits 0, 2, 4, 6
	\end{minted}
	\caption{Basic RxJS example creating an Observable emitting four integers. Each integer is processed by two operators and finally written to the console.}
	\label{lst:rp-with-rxjs}
\end{listing}

@@

\begin{figure}[H]
	\centering
	\tikzfig{marble-diagram}
	\caption{A Marble Diagram visualizing the Observable in Listing~\ref{lst:rp-with-rxjs}. From left to right, each marble represents an emitted value. The vertical line at the last marble indicates that the Observable completed after emitting \mintinline{TypeScript}{6}.}
	\label{fig:marble-diagram}
\end{figure}
